---
title: "Fossil CO2 Emissions"
format: 
  html:
    toc: true
    embed-resources: true
editor: visual
---

```{r}
#| include: false
library(tidyverse)
library(lubridate)
library(tbl2xts)
library(dygraphs)
library(pier)
library(ggplot2)
library(plotly)
library(scales)
library(treemapify)
library(paletteer)
library(treemap)
library(readxl) 
library(remotes) 
library(tidyverse) 
library(ggalt) 
library(usmap) 
library(ggplot2) 
library(vtable) 
library(maps) 
library(lubridate) 
library(stringr)
library(scales)
library(stats)
library(leaflet)
library(ggrepel)
library(RColorBrewer)
library(gghighlight)
library(treemap)
library(gridExtra)
library(readr)
library(mapproj)
library(scales)
library(plotly)
library(RColorBrewer)
library('unikn') 

# Viz 1
Final_Project_data <- read_csv("List of countries by carbon dioxide emissions - Copy.csv")
# create world map data
# create world map data
world_map <- map_data("world") 
# join data with world map data
data_map <- Final_Project_data %>% left_join(world_map, by = c("Countries" = "region"))
data_map_nona <- data_map %>% filter(!is.na(lat) & !is.na(long))

# filter data for top 10 countries
# filter data for top 10 countries
data_top10 <- Final_Project_data %>%
  filter(!grepl("World", Countries)) %>%
  arrange(desc(`Fossil CO2 emissions(Mt CO2) 1990`)) %>%
  slice_max(n = 10, order_by = `Fossil CO2 emissions(Mt CO2) 1990`)
# reorder factor levels of Countries in decreasing order of CO2
data_top10$Countries <- fct_reorder(data_top10$Countries, data_top10$`Fossil CO2 emissions(Mt CO2) 1990`, .desc = TRUE)

# Viz 2
# create world map data
world_map1 <- map_data("world")
# join data with world map data
data_map1 <- Final_Project_data %>% left_join(world_map1, by = c("Countries" = "region"))
data_map_nona1 <- data_map1 %>% filter(!is.na(lat) & !is.na(long))

# filter data for top 10 countries
# filter data for top 10 countries
data_top101 <- Final_Project_data %>%
  filter(!grepl("World", Countries)) %>%
  arrange(desc(`Fossil CO2 emissions(Mt CO2) 2005`)) %>%
  slice_max(n = 10, order_by = `Fossil CO2 emissions(Mt CO2) 2005`)
# reorder factor levels of Countries in decreasing order of CO2
data_top101$Countries <- fct_reorder(data_top101$Countries, data_top101$`Fossil CO2 emissions(Mt CO2) 2005`, .desc = TRUE)

# Viz 3
# create world map data
world_map2 <- map_data("world")
# join data with world map data
data_map2 <- Final_Project_data %>% left_join(world_map2, by = c("Countries" = "region"))
data_map_nona2 <- data_map %>% filter(!is.na(lat) & !is.na(long))

# filter data for top 10 countries
# filter data for top 10 countries
data_top102 <- Final_Project_data %>%
  filter(!grepl("World", Countries)) %>%
  arrange(desc(`Fossil CO2 emissions(Mt CO2) 1990`)) %>%
  slice_max(n = 10, order_by = `Fossil CO2 emissions(Mt CO2) 2017`)
# reorder factor levels of Countries in decreasing order of CO2
data_top102$Countries <- fct_reorder(data_top102$Countries, data_top102$`Fossil CO2 emissions(Mt CO2) 2017`, .desc = TRUE)

# Viz 4
 t <- data_map %>% filter(!grepl("World", Countries)) %>% select(Countries,`Fossil CO2 emissions(Mt CO2) 2017`,`Fossil CO2 emissions(Mt CO2) 1990`, `Fossil CO2 emissions(Mt CO2) 2005`) %>% 
  pivot_longer(cols = -Countries, names_to = "Year_m", values_to = "CO2 Emissions") %>%
  mutate(Year = case_when(Year_m == "Fossil CO2 emissions(Mt CO2) 1990" ~ 1990,
Year_m == 'Fossil CO2 emissions(Mt CO2) 2005' ~ 2005,
Year_m == 'Fossil CO2 emissions(Mt CO2) 2017' ~ 2017))
 
# Viz 5
# load the data
stateco2 <- read_xlsx('State energy-related CO2 Emissions by year.xlsx')
# load prof's package
library(remotes)
remotes::install_github('NickCH-K/nicksshorts')
help(package = 'nicksshorts')
library(nicksshorts)
# Clean the data in excel file first: I deleted 1st – 4th and last 3 (57th –59th) rows, and only keep the column of 2020 
# convert US state name to abbreviations
stateco2 <- stateco2 %>% mutate(State_abb = state.abb[match(State,state.name)]) %>% 
  relocate(State_abb, .after = State)
stateco2[is.na(stateco2)] <- "DC"
# rename the colname
colnames(stateco2)[1] ="State_name"
colnames(stateco2)[2] ="state"
```

Picture this: it's the middle of summer, and you're walking down a city street. The air is thick and muggy, and you can feel the weight of it on your chest. The smell of exhaust fumes and other pollutants fills your nostrils, making it difficult to breathe. This is just one small example of the harmful effects of fossil CO2 emissions on our planet. The burning of fossil fuels such as coal, oil, and gas has led to a dangerous buildup of carbon dioxide in the atmosphere, causing global temperatures to rise and wreaking havoc on our climate and ecosystems. The dangers of fossil CO2 emissions are numerous, and it is imperative that we take immediate action to reduce them before it is too late.

We performed an analysis to analyze the trends, and the percentage changes in carbon dioxide emissions over time (1990, 2005, 2017) for different countries, and identify the countries with the highest and lowest emissions compared to the global standards. This paper addresses the results of our analysis on the International Fossil CO2 Emissions, but, in particular, the Fossil CO2 Emissions in United States, identifies the major sectors and the type of fuel producing the most emissions. The final part of this paper focuses on the major insights, and proving recommendations to minimize the carbon dioxide emissions in those sectors.

#### International Fossil CO2 Emissions

The heatmap and the bar plots show the global carbon emissions in three decades - 1990, 2005, and 2017.

```{r}
#| echo: false
#| warning: false
#| message: false
# create a world map with CO2 emission values
p <- ggplot(data_map_nona, aes(x = long, y = lat, group = group, fill = `Fossil CO2 emissions(Mt CO2) 1990`)) +
  geom_polygon() +
  scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), guide = guide_colorbar(barwidth = 20, barheight = 0.5)) +
  labs(x = "", y = "",fill="Fossil CO2 emissions\n(Mt CO2) " ,title = "International Fossil CO2 Emissions in 1990") +
  theme(
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.line = NULL )
ggplotly(p, tooltip = c("group", "Fossil CO2 emissions(Mt CO2) 1990"), labels = c(format(x = scales::comma, big.mark = ",", accuracy = 0), "Mt CO2")) 
```

```{r figg, fig.align = 'center', fig.height = 3, fig.width = 3.5}
#| echo: false
#| warning: false
#| message: false
ggplot(data_top10, aes(x = Countries, y = `Fossil CO2 emissions(Mt CO2) 1990`)) +
  geom_col(fill = "red") +
  labs(x = " ", y = " ", 
       title = "Highest Emissions (Million Tonnes)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
     theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    theme(axis.line = element_line(color = 'black'), 
          axis.ticks.x = element_line(size = 5), 
          axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"),
          panel.background = element_rect(fill = 'white'), 
          plot.title = element_text(hjust = 0.5)) +
    gghighlight(Countries %in% c('USA'),
                unhighlighted_params = aes(size=.1), 
                label_params=list(direction='y',nudge_x=10)) 
```

In 1990, United States produced the most CO2 Emissions, with the European Union and China following them respectively. Back in the days, the economy of United States was largely dependent on manufacturing, while the Tech industry was just beginning to take shape.

```{r}
#| echo: false
#| warning: false
#| message: false
# create a world map with CO2 emission values
p1 <- ggplot(data_map_nona1, aes(x = long, y = lat, group = group, fill = `Fossil CO2 emissions(Mt CO2) 2005`)) +
  geom_polygon() +
  scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), guide = guide_colorbar(barwidth = 20, barheight = 0.5)) +
  labs(x = "", y = "",fill="Fossil CO2 emissions\n(Mt CO2) " , title = "International Fossil CO2 Emissions in 2005") +
  theme(
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.line = NULL )
ggplotly(p1, tooltip = c("group", "Fossil CO2 emissions(Mt CO2) 2005"), labels = c(format(x = scales::comma, big.mark = ",", accuracy = 0), "Mt CO2")) 
```

```{r figgg, fig.align = 'center', fig.height = 3, fig.width = 3.5}
#| echo: false
#| warning: false
#| message: false
# plot bar chart
ggplot(data_top101, aes(x = Countries, y = `Fossil CO2 emissions(Mt CO2) 2005`)) +
  geom_col(fill = "red") +
  labs(x = " ", y = " ", 
       title = "Highest Emissions (Million Tonnes)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    theme(axis.line = element_line(color = 'black'), 
          axis.ticks.x = element_line(size = 5), 
          axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"),
          panel.background = element_rect(fill = 'white'), 
          plot.title = element_text(hjust = 0.5)) +
    gghighlight(Countries %in% c('USA'),
                unhighlighted_params = aes(size=.1), 
                label_params=list(direction='y',nudge_x=10))
```

Moving a decade ahead, United States has maintained its carbon emission levels without any significant increase or decrease; however, China, owing to great surge in manufacturing and construction activities, significantly surpassed all other regions. Although the CO2 emissions from United States had been constant between these years, those numbers are high i.e., United States had been constantly producing high CO2 Emissions, which is sublimed by China's even higher numbers.

```{r}
#| echo: false
#| warning: false
#| message: false
# create a world map with CO2 emission values
p2 <- ggplot(data_map_nona2, aes(x = long, y = lat, group = group, fill = `Fossil CO2 emissions(Mt CO2) 2017`)) +
  geom_polygon() +
  scale_fill_gradientn(colours = brewer.pal(9, "YlOrRd"), guide = guide_colorbar(barwidth = 20, barheight = 0.5)) +
  labs(x = "", y = "", fill="Fossil CO2 emissions\n(Mt CO2) " ,title = "International Fossil CO2 Emissions in 2017") +
  theme(
        panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.background = element_blank(),
        panel.line = NULL )
ggplotly(p2, tooltip = c("group", "Fossil CO2 emissions(Mt CO2) 2017"), labels = c(format(x = scales::comma, big.mark = ",", accuracy = 0), "Mt CO2")) 
```

```{r figggg, fig.align = 'center', fig.height = 3, fig.width = 3.5}
#| echo: false
#| warning: false
#| message: false
# plot bar chart
ggplot(data_top102, aes(x = Countries, y = `Fossil CO2 emissions(Mt CO2) 2017`)) +
  geom_col(fill = "red") +
  labs(x = " ", y = " ", 
       title = "Highest Emissions (Million Tonnes)") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    theme(axis.line = element_line(color = 'black'), 
          axis.ticks.x = element_line(size = 5), 
          axis.text.x = element_text(face = "bold"),
        axis.text.y = element_text(face = "bold"),
          panel.background = element_rect(fill = 'white'), 
          plot.title = element_text(hjust = 0.5)) +
    gghighlight(Countries %in% c('USA'),
                unhighlighted_params = aes(size=.1), 
                label_params=list(direction='y',nudge_x=10)) 
```

In the next decade, the increase in China's CO2 emissions makes the emissions from United States looks negligible. Yes, it has reduced from 6000 mega tonnes of CO2 released in 2005, to 5000 mega tonnes in 2017. But they are not low values when compared to the global standards!

This sets the stage for further analysis of United States' Fossil CO2 Emissions and the factors contributing to them.

#### United States Fossil CO2 Emissions

```{r}
#| echo: false
#| warning: false
#| message: false
t %>% ggplot(aes(x = Year, y = `CO2 Emissions`, group = Countries, color = Countries)) +
  geom_line(size = 1.5, show.legend = TRUE) +
  scale_color_manual(values = c("USA" = "red", "China" = "blue")) +
  labs(x = " ", y = "CO2 Emissions in Mt", title = "CO2 Emissions: United States vs China") +
  scale_x_continuous(limits = c(1990, 2017), breaks = c(1990, 2005, 2017)) +
  gghighlight(Countries %in% c("USA", "China"),
              unhighlighted_params = aes(size = 0.1),
              label_params = list(direction = "y", nudge_x = 10, hjust = 0)) +
  geom_text(data = filter(t, Countries == "USA" & Year == 2017),
            aes(x = Year, y = `CO2 Emissions`, label = "USA"),
            nudge_y = 20, color = "red") +
  geom_text(data = filter(t, Countries == "China" & Year == 2017),
            aes(x = Year, y = `CO2 Emissions`, label = "China"),
            nudge_y = -20, color = "blue") +
  theme_minimal(base_family = "serif") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
```

This line graph compares the Fossil CO~2~ Emissions in United States and China between the years 1990 and 2017. United States had a 0.40% change between 1990 and 2017, which shows emissions remained constant over nearly three decades. However, for China's emissions increased by 353.80%, and total world emissions increased by 63.50%. Though, China had the highest emissions in 2017, look at the gradual decrease in the emissions from United States. How can we reduce these high numbers further?

Look at this!

```{r}
#| echo: false
#| warning: false
#| message: false
plot_usmap(data = stateco2[,2:3], regions = "states", values = "2020") + 
  scale_fill_continuous(low = "white", high = "red", name = "CO2 Emissions\n(Mt CO2)") + 
  theme(legend.position = "right") + 
  labs(title = "United States Fossil CO2 Emissions in 2020")
```

Doesn't the state with the bright red color standout? Yes, that's Texas, with the highest CO2 emissions in the United States.

```{r}
#| echo: false
#| warning: false
#| message: false
stateco2$`2020` <- format(stateco2$`2020`, decimal.mark = ".", digits = 2)
stateco2$hover <- with(stateco2, paste(State_name, '<br>', "CO2 Emissions:", `2020`))
g <- list(
  scope = 'usa',
  projection = list(type = 'albers usa'),
  showlakes = TRUE,
  lakecolor = toRGB('white')
)
plot_geo(stateco2, locationmode = 'USA-states') %>% add_trace(
  z = stateco2$`2020`, text = ~hover, locations = ~state, 
  color = ~2020, colorscale = 'Red') %>%  
  colorbar(title = "CO2 Emissions\n(Mt CO2)") %>% 
  layout(
    title = "United States Fossil CO2 Emissions in 2020",
    geo = g
  )
```

From this interactive visual, we can conclude that the states with the highest emission is Texas. The carbon emissions from Texas are twice as high as that from California, the state with the second highest carbon footprint in USA. Why are the emissions so high in Texas? Texas is a big city with many manufacturing and technological industries. How can it be reduced?

Let's divide the emissions by sectors and the type of fuel producing those emissions.

```{r}
#| echo: false
#| warning: false
#| message: false
#Visualization 6
b <- tibble(Sector = c("Industrial","Transportation","Electric Power","Commercial","Residential"),
               Emission = c(226.2,192.7,176.9,8.1,9.1))
treemap(b,
       index=c("Sector"),
       vSize="Emission",
       vColor="Emission",
       type="value",
       palette=c("#FFEFEF","pink","indianred"),
       fontcolor.labels = "black",
  fontface.labels = c("bold", rep("plain")),
  fontfamily.title = "sans",
  fontfamily.labels = "sans",
  fontfamily.legend = "sans",
  border.col="white",
       title.legend="CO2 Emissions (Million Tonnes)",
       title = "CO2 Emissions by Sector",
       format.legend = list(scientific = FALSE, big.mark = " "))
```

To provide sector-specific suggestions, the emissions are divided into five sectors. High and comparatively similar amounts of CO2 is produced in three sectors - Industrial, Transportation and Electric Power. Carbon emissions in the industrial sector comes from the energy required to manufacture goods, operate machines, etc. In the transportation sector, CO2 is released from the burning of fuels like petroleum and gas required to operate vehicles.

```{r}
#| echo: false
#| warning: false
#| message: false
# Visualization 7
data <- data.frame(label = c('Coal','Natural Gas','Petroleum'),
                   value = c(83.6,243.9,296.5),
                   color = c("Coal"='pink',"Petroleum"='salmon',"Natural Gas"='indianred'))
advanced.pie <- data %>%
    pier() %>%
    pie.size(inner=70, outer=100, width = 600, height = 450) %>%
    pie.header(text='Texas CO2 Emissions', font='Impact', location='pie-center') %>%
    pie.subtitle(text='by Fuel') %>%
  pie.tooltips()%>%
   pie.labels(mainLabel = list(font = 'Impact',
                              fontSize = 14))
htmlwidgets::onRender(advanced.pie, 'document.getElementsByTagName("svg")[0].setAttribute("viewBox", "")')
```

In the modern era, usage of coal for producing energy is not as common as using petroleum and natural gas. In Texas, high amounts of carbon is released from activities involving the combustion and usage of petroleum and natural gas.

```{r}
#| echo: false
#| warning: false
#| message: false
# Visualization 8
data <- tibble(Sector = c(rep("Industrial \n Sector",3),rep("Transportation \n Sector",3),rep("Electric Power",3)),
               Fuel = c('Coal','Petroleum','Natural Gas','Coal','Petroleum','Natural Gas','Coal','Petroleum','Natural Gas'),
               Emission = c(3.0,109.3,119.3,3.0,182.4,15.2,83.0,3.0,93.8))
data$Sector <- factor(data$Sector,levels = c("Transportation \n Sector", "Industrial \n Sector", "Electric Power", "Commercial","Residential"))
finall<-ggplot(data, aes(x = Sector, y = Emission,fill=Fuel)) + 
  geom_col(position = 'dodge')+
  theme_light()+
    theme(
    panel.grid.major.x = element_blank(),
    legend.position = "bottom"
  )+
      labs(x = ' ', y = 'CO2 Emissions (Million Tonnes)',fill='', title = "CO2 Emissions by Sector and Fuel") +
   theme(axis.text.x  = element_text(face = 'bold'))+
  theme(axis.text.x  = element_text(size=10))+
    scale_fill_brewer(palette = "OrRd")
ggplotly(finall)
```

#### Conclusion

From the graph, we can see that the emissions from coal to be minimal compared to the others. Natural gas usage is high in the Industrial Sector, and used to the produce electrical power. Another important point to consider from this visual is that petroleum is one of the major producers of CO2.

From these observations, the most obvious thing to do to reduce the emissions from the transportation sector is to switch to Electric Vehicles. Electric vehicles are booming now and that's good. Further, we can encourage people to use public transportation systems like buses, trains, and subways to reduce the number of personal vehicles on the road, which in turn reduces emissions. The Governments can mandate fuel efficiency standards for vehicles, which can lead to reduced emissions and lower fuel costs for consumers. We all know that a high amount of carbon is produced in the process of generating electricity. It's obvious! Switch to Solar Power folks. Also, the industries can adapt to using solar power to run the machines.

Reducing fossil CO2 emissions is critical for the long-term well-being of both humans and the planet, and should be a top priority for individuals, businesses, and governments. So yeah! Let us all consider these solutions and be less toxic to the atmosphere.
